---
title: "Covid Clinics in Toronto"
title-block-banner: true
#subtitle: "TBD"
abstract: "TBD"
thanks: "Code and data supporting this analysis is available at: https://github.com/christina-wei/INF3014-1-Covid-Clinics.git"
author: "Christina Wei"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: ref.bib
---

```{r}
#| message: false
#| echo: false

#### Workspace set-up ####

## Load packages
library(tidyverse) # a collection of data-related packages

#### Read in cleaned data ####
cleaned_data = read_csv(
  file = "../inputs/data/cleaned_clinic_data_combined.csv",
  show_col_types = FALSE
)
```

## Introduction

-   Overall Covid-19 situation worldwide

-   How Canada handled Covid cases

-   Sustained immunization effortin Canada today - dataset

-   Potential biases to watch out for - marginization

## Data

Overview of the data

-   What datasets are used

-   What type of data are included

-   What R packages are used in analysis

Description of the cleaning process

Data cleaned by using the open source statistically programming language R [@r]. Tidyverse reference [@rTidyverse], which includes packages ggplot2 [@rGgplot2], dplyr [@rDplyr], readr [@rReadr], tibble [@rTibble] (https://www.tidyverse.org/packages/). Use janitor to clean names \[@rJanitor\]

Data downloaded using [@rOpenDataToronto], data sources from \[@torontoCityWards\], [@torontoCovidClinics], [@torontoWardProfiles]

Measurements

-   Doesn't not track clinic openings and closing over time

-   Hours and time are not tracked properly - just because there is a clinic doesn't mean it is accessible - eligibility criteria, hours etc.

-   Population from 2018 - a bit outdated

-   Would be useful to have utilization rates as well

## Results

Clinics added per month

Clinic per district

Clinic by type

```{r}
#| echo: false

cleaned_data |>
  ggplot(aes(x = ward_name, fill=location_type)) + 
  geom_bar() + #bar graph
  theme_minimal() + #neater theme
  theme(axis.text.x = element_text(angle=90)) +
  labs(
    x = "Party",
    y = "Number of seats"
  ) #add labels
```

## 

## Discussion

## Conclusion

## Appendix

## Reference
